 

### 2021-05-24 vertical-thermal-review.Rmd
work on T_leaf

**The biophysical drivers reviewed above influence leaf temperature $T_{leaf}$ distribution along the vertical gradient.**
Along the biophysical gradients, leaves adjust to their environment by maintaining, when possible, a $T_{leaf}$ that is optimal for carbon assimilation and metabolic processes  [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy). 
Often, however, $T_{leaf}$ remains similar to $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016 @michaletzPlantThermoregulationEnergetics2015].
This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. During $gs$, evapotranspiration facilitates heat dissipation. But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].

@@

High wind speed (Fig. `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a]. Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]

@@

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to smaller $ga$ than larger leaves that take longer due to large $ga$. Therefore, $T_{leaf}$-$T_{air}$ is lower for small leaves than large leaves that can exprience greater $T_{leaf}$-$T_{air}$. With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures than large leaf species [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

>>>>

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, a that is optimal for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy).
Often, however, $T_{leaf}$ tracks $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020]. 
Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost (such as transpiration) to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].

@@

*(In general, we'll want to adjust this section to parallel the figure and tie in with the section above.)*

@@

High wind speed (Fig. `r fig_NEON_vertical`) and large $g_a$ decrease $T_{leaf}$ - $T_{air}$ [Fig. `r fig_leaf_T`b, @daudetWindSpeedLeaf1999a]. 
Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]

@@

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to smaller $g_a$ than larger leaves that take longer due to large $ga$. Therefore, $T_{leaf}$-$T_{air}$ is lower for small leaves than large leaves that can exprience greater $T_{leaf}$-$T_{air}$. With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures than large leaf species [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. 
When stomatal conductance ($g_s$) is high, evapotranspiration facilitates heat dissipation. 
However, when high irradiance, $T_{air}$, $VPD$ or drought cause leaf water demand to exceed water transport, conservation of water ($g_s$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].
Therefore, while under maximum $g_s$, leaves can maintain $T_{leaf}$ closer to $T_{air}$ at high solar-radiation loads until light saturation, during $g_s$ limitation solar radiation can drastically elevate $T_{leaf}$ above $T_{air}$, especially for larger leaves [@fauset_differences_2018; @songTreeSurfaceTemperature2020; @pauTropicalForestTemperature2018].
